A modelling strategy to improve cacao quality and productivity

### 2021-11-05 paperModel/paperModel.tex
Add files via upload

% Authors, for the paper (add full first names)
\Author{Angela Romero V $^{1,\ddagger}$*\orcidA{}, Adriana Gallego $^{2}$ and Anyela V. Camargo R.$^{1, \ddagger}$}

@@

\section{Introduction}
Cocoa ( \textit{Theobroma cacao }L. ) is an important worldwide perennial tropical crop endemic to the South American rainforests \citep{zuidema2005, motamayor2002, argout2011, Rodriguez2019}. Cacao plant member of the Malvaceae (formerly Sterculiaceae)  botanical family such as  cotton \textit{ Gossypium hirstium} \citep{Nix2017cotton} wich is modeled in SIMPLE model \citep{Zao2019simple}. Cocoa is grown for its fruits, known as cacao pods.  \citep{ Niemenak2010, suarez2021}. Only the 5\% of the world cocoa yield is desalinated for Fine-cocoa production due to the low productivity of  the traditional crop management \citep{argout2011}.  In Colombia, cocoa  is  traditionally  consumed  as  a  beverage. It is one of the crops promoted by the Colombian government in the social and agricultural development  programs aimed at favouring peace in post-conflict regions \citep{Rodriguez2019, Abbott2019}. This crop is grown by approximately 52.000 families \citep{Gutierrez2020} and 98\% of production being carried out by small and medium-sized producers \citep{Garcia2014, Escobar2020}. Colombia registered an increase of 3.750 tons in production in 2020 compared to the previous year \citep{lamos2020}.

Although Colombian cocoa has the potential to be in the high value markets for fine flavour \citep{Escobar2020}, it is still not widely produced as the lack of adoptions of technologies by the traditional farmers. They empirically harvest after 5 or 6 months after flowering date, hence they ferment cocoa beans without considering the  quality of the seeds at the harvest time. This produce heterogeneous characteristics between each fermentation batch diminishing the quality of cocoa final product \citep{Escobar2021}. To identify the best moment to harvest is important to consider physiological responses affected by climate variables such as rain, solar radiation and wind.  Thus,  for cocoa in Colombia, physiological simulation models may be valuable to identify the best moment to harvest cocoa considering variable weather conditions, soil types and cultivar specifications.

@@

\begin{figure}[h!]
	\centering
	\includegraphics[scale=0.4]{images/phenology.png}\\
	\caption{\footnotesize {Phenology of cocoa in Colombia for crop modelling.\\}} 
	\footnotesize{Credits: Taken from from Dreamstime.com, phys.org \citep{toledo2021} and \cite{lopez2018}}
	\label{fig:pheno}
\end{figure}

@@

\subsection{Weather conditions}
\begin{figure}[h!]
	\centering
	\includegraphics[scale=0.4]{images/clima.png}
	\caption{\footnotesize {Colombian weather conditions. \\}}	
	\label{fig:temp}
	{\footnotesize (a), available photosynthetic solar radiation (PAR). (b), Daily average temperature  (c), Precipitation and relative humidity per month}
\end{figure}
\newpage

@@

Calibration of crop models are conducted typically for particular cultivars and require site specific inputs for weather \citep{Crout20142}. The procedure for the SIMPLE model \citep{Zao2019simple} calibration  was a sequential process of modifying physiological variables specific for cocoa in the inputs files and  adding the appropriate files of weather for each region tested. In the SIMPLE model has seven input files where new cocoa crop data should be provided. Files in the list below can be edited to define the features of the new cultivars or experiments. Cocoa crop has not been simulated with SIMPLE model, hence consider previous cocoa studies, values such as leaf area index (LAI) \citep{Agele2016} and Harvest index (HI) \citep{Quintana2015} where modified. Radiation Use Efficiency (RUE) was calibrate according to the perennial crops  (banana and cotton) that had been calibrated previously in SIMPLE model \citep{Zao2019simple} with a RUE of 0.8 and 0.85 respectively. As cocoa trees are under shadow the RUE was lower with values between 0.7 and 0.5 g MJ$^{\mathbf{-1}}$ m$^{2}$ (table \ref{tab:reparam}).
Once the physiological parameters were calibrated to the  simulated yields for cocoa were reasonably close to the observed yield. 23 flowering dates from 12-July-2019 to 23-June-2020 were introduced in the treatment file, running the program and saving the results.

@@

\subsection{Thermal time }
Thermal time characterization is shown in figure \ref{fig:ttbox}. It showed differences per region as was expected following the tendencies of the temperature per region (fig. \ref{fig:temp},c). Apartado and Arauca had the highest temperatures hence, the highest thermal time values with 2909 and 2764 days$^\circ$C  respectively. Caldas had the lowest values with 1173 days$^\circ$C. Meanwhile, Cali and Santander presented similar thermal time around 2000 $^\circ$C (table \ref{tab:reparam}). The accumulated temperature during the pod development (fig.\ref{fig:ttbox}) depends on the region where cocoa is cultivated and the variety planted. Thermal time values are also proportional to the yield reported on field (table \ref{tab:reparam}).\\

\begin{figure}[h!]
	\centering
	\includegraphics[scale=0.3]{images/ttbbox.png}\\
	\caption{\footnotesize {Coaoa yield and thermal time characterization.\\}}
	\label{fig:ttbox}
\end{figure}

@@

\begin{figure}[h!]
	\centering
	\includegraphics[scale=0.35]{images/heatm.png}\\
	\caption{\footnotesize {Pearson correlation among weather variables and flowering dates all locations. \\}}
	\label{fig:heat} 
\end{figure}

\subsection{Model validation}
The cocoa yield simulation (kg ha$^{-1}$ per year) was validated using observed data (table \ref{tab:reparam}) showing a final RRMSE of 7.2\% (fig. \ref{fig:m1}, a). Individual errors per region are presented in table \ref{tab:error}, where the best fit simulation model was for yield prediction in Apartado and the highest error was calculated for Caldas crops.  The model responded to the variations of temperature and solar radiation. Therefore, the highest  yield values simulated were obtained for Arauca over 4000 kg ha$^{-1}$, followed by Apartado Santander with yields over 2000 kg ha$^{-1}$.  The lowest yield was simulated for Caldas region with less of 1000 kg ha$^{-1}$ . Final yield in the model was is calculated as the product of biomass of aerial part and harvest index (HI) \citep{Zao2019simple, Amir1991}, where the HI is similar to the CropSyst \citep{STOCKLE2003} and AquaCrop \cite{Steduto2009}.

Therefore, biomass production of the aerial part of the plant was simulated per region (fig. \ref{fig:m1}, b) as biomass-rate which is the daily biomass growth rate.  As for cocoa simulation we did not consider water and heat stress variables, in this research biomass simulation is affected by the solar radiation, RUE, daily temperature, atmospheric CO$_{2}$ concentration (ppm) and the fraction of solar radiation intercepted by a tree of cocoa during the fruit development (fSolar) (fig. \ref{fig:m1}, c). Due to the lack of this kind of data from fields, the comparison was not possible with observed data of biomass production through the crop cycle. However, the biomass simulation also responded proportionally with the yield production. Biomass of cocoa fruits simulated include all the parts that compound  the pod when they are fresh. Therefore, these this information may be of interest to farmers to predict pod fresh mass.

The figure \ref{fig:m1}, c the fSolar was simulated thought the fruit development resulting that crops in  all regions reached the maximun fSolar at 0.94, except Caldas where crops reached 0.76.  Apartado and Arauca reached faster this fSolar-max. These high values of solar radiation intercepted for photosynthesis, lasted differently depending on the region and their solar radiation (Fig.\ref{fig:temp}, a) : Apartado 66 days  from  69 to 135 days after flowering (daf),  Arauca 61 days from 72 to 133 daf, Cali 38 days from 83 to 121 daf,  Santander 24 days from 92 to 116 daf and Caldas 2 days at 125 daf. fSolar declined in the interception of solar radiation until the pod harvest day.

\begin{figure}[h!]
	\centering
	\includegraphics[scale=0.4]{images/outmodel.png}
	\caption{\footnotesize {SIMPLE model output for cocoa in Colombia.\\ }} 
	\label{fig:m1}
\end{figure}

\begin{table}[h!]	
	\caption {\footnotesize {Summary of relative root mean square error RMSE for the predicted cocoa yield using SIMPLE model.}}
	\label{tab:error} 
	\centering
	\begin{small}
		\begin{tabular}{l c c c c c c}
			\hline
			{\bf Region }&{\bf Apartado }&{\bf Arauca}&{\bf Santander}&{\bf Cali}&{\bf Caldas}&{\bf Overal}\\
			\hline
			RMMSE \%  & 3 & 6.05 & 10.06&8.5&14.90&7.2 \\
			\hline
		\end{tabular} \\
	\end{small}
\end{table}

\subsection{Prediction of the pod harvest day }
These result demonstrated that the traditional way to harvest always at 180 daf is not having account physiological and environmental conditions that can be affecting the pod development and maturation.
Figure \ref{fig:dayH} present that observe day of harvest or maturity day is 180 daf independently of the region. All the regions except Santander, presented the earliest harvest when the flowering date was between December and February. The fruit can be ready to harvest depending of the enviromental condition per region. Thus, when the fruit development was simulated the maturity day in Cali and Caldas were from 3 to 12 and 4 to 23 days before 180 daf (Fig. \ref{fig:dayH}, c and f respectively). Apartado presented the most similar predicted dates of harvest to 180 daf with 170 to 182 daf  (Fig. \ref{fig:dayH}, d). The pod may be harvest in Arauca (Fig.\ref{fig:dayH}, b)  between 165 and 193 daf, Santander (Fig.\ref{fig:dayH}, a) between 165 and 183 daf, ten days less than in Arauca for the same months of flowering of July and August of 2019 .

\begin{figure}[h!]
	\centering
	\includegraphics[scale=0.4]{images/RegionHarvest2.png}
	\caption{\footnotesize {Day of harvest cocoa simulation\\ }} 
	\label{fig:dayH}
\end{figure}.
\newpage

@@

\section{Discussion}
Authors should discuss the results and how they can be interpreted from the perspective of previous studies and of the working hypotheses. The findings and their implications should be discussed in the broadest context possible. Future research directions may also be highlighted.

This study presents a new approach of SIMPLE model calibration for cocoa as tropical crop in South America for five environments.

The optimum temperature for photosynthesis in cacao has
been reported to be between 31 °C and 33 °C (Balasimha et al.
1991) and 33 °C–35 °C (Yapp 1992). Sena Gomes and
Kozlowski (1987) reported a decrease in gs over the range
18.7 °C to 27.2 °C and an increase at temperatures above this,
most likely to enhance cooling at higher temperatures.
However, the opposite effect was described by Raja Harun
and Hardwick (1988a). In this study, gs and transpiration (E)
increased within the range of 20 °C–30 °C. In field-grown
cacao in India, photosynthetic rate declined as mean monthly
temperature increased above 34 °C during the dry season
(Balasimha et al. 1991). \cite{lahive2019}
\textcolor{red}{this section on going to organize}

Evaluating the level of knowledge of producers regarding cocoa crop management, the harvest was in the group of activities that presented the lowest level of knowledge on the part of the producers according to the general averages \citep{Gutierrez2020}.

\subsection{Weather Effects}
Wind can affect the availability of tiny flies pollinators from Diptera order and from the families of of the biting midges \textit{Ceratopogonidae},  genus  \textit{Forcipomyia} \citep{Saunders1959, kaufmann1975, sotomayor2020} to reach the cocoa flowers. However, wind is not included in the SIMPLE model as an input as cocoa crop has not been simulated with this model before. Flower opening is very well synchronised between the cohorts of mature flowers opening each night. The flowers open at almost exactly the same time and rate, irrespective of their position on the trunk. Unfertilised flowers abscise from the trunk approximately 1 day after flower opening  \citep{Niemenak2010}.  In a preliminary analysis of Santanter data, We found that the number of successful flowers pollinated to produce final yield can be affected mostly by the rain, TMAX and wind (fig.\ref{fig:heat}).

\subsection{App development and future challenges }
The original code of SIMPLE model of \citep{Zao2019simple} was modified to make easy the implementation of this cocoa model as and app to be used in smartphones and desktops by farmers in Colombia. Therefore, the new version for cocoa crops simulation will be used to predict yield, date of harvest and biomass production, inserting only the date of flowering and region. The app development is on charge of Grupo BIOS to be deliver to farmers in Caldas initially at the end of 2021. 
This research presented and initial crop calibration that can be improved with further studies, including effects over the pod production by diseases, nutritional diffidences and abiotic stresses. The future challenge will be that traditional farmers start to harvest more aware of the environmental effects over their crops. It will be necessary that they engage growers with adapt founding from scientific studies. Moreover, It will be required the help of entrepreneurs, researchers, academics and non-specialized communities to transfers the knowledge to cocoa growers.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
Based on principles of crop physiology this  generic crop model , which was developed with relatively few equations and parameters was fitted for cocoa fruit growth. The SIMPLE model worked well for cocoa yield prediction and to demonstrate that the moment to harvest the cocoa pod is affected by the weather variations over the physiology of the trees.

\acknowledgments{The authors wish to thank to Grupo BIOS and Fedecacao for facilitated the data from cocoa fields. In this section you can acknowledge any support given which is not covered by the author contribution or funding sections. This may include administrative and technical support, or donations in kind (e.g., materials used for experiments).}

>>>>

% Authors, for the paper (add full first names)
\Author{Angela Romero V $^{1,\ddagger}$*\orcidA{}, Anyela V. Camargo R. $^{1, \ddagger}$ and Adriana M. Gallego$^{2}$}

@@

\section{Introduction}
Cocoa ( \textit{Theobroma cacao }L. ) is an important worldwide perennial tropical crop endemic to the South American rainforests \citep{zuidema2005, motamayor2002, argout2011, Rodriguez2019}. Cacao plant is a member of the Malvaceae (formerly Sterculiaceae)  botanical family such as  cotton \textit{ Gossypium hirstium} \citep{Nix2017cotton}. Coton has been is modeled in SIMPLE model \citep{Zao2019simple}. Cocoa is grown for its fruits, known as cacao pods \citep{ Niemenak2010, suarez2021}. Only the 5\% of the world cocoa yield is destined for Fine-cocoa production due to the low productivity through  the traditional crop management \citep{argout2011}.  In Colombia, cocoa  is  traditionally  consumed  as  a  beverage. It is one of the crops promoted by the Colombian government in the social and agricultural development  programs aimed at favouring peace in post-conflict regions \citep{Rodriguez2019, Abbott2019}. This crop is grown by approximately 52.000 families \citep{Gutierrez2020} and 98\% of production being carried out by small and medium-sized producers \citep{Garcia2014, Escobar2020}. Colombia registered an increase of 3.750 tons in production in 2020 compared to the previous year \citep{lamos2020}.

Although Colombian cocoa has the potential to be in the high value markets for fine flavour \citep{Escobar2020}, it is still not widely produced as the lack of adoptions of technologies by the traditional farmers. They empirically harvest at 5 or 6 months after flowering date, hence   a mix of quality of cocoa beans can be fermented. This produce heterogeneous characteristics between each fermentation batch diminishing the quality of cocoa final product \citep{Escobar2021}. To identify the best moment to harvest is important to consider physiological responses affected by climate variables such as rain, solar radiation and wind.  Thus,  for cocoa in Colombia, physiological simulation models may be valuable to identify the best moment to harvest cocoa considering variable weather conditions, soil types and cultivar specifications.

@@

\begin{figure}[h!]
	\centering
	\includegraphics[scale=0.4]{images/phenology.png}\\
	\caption{\footnotesize {Phenology of cocoa in Colombia for crop modelling.\\}} 
	\footnotesize{Source: Taken from from Dreamstime.com, phys.org \citep{toledo2021} and \cite{lopez2018}}
	\label{fig:pheno}
\end{figure}

@@

\subsection{Weather conditions}
\begin{figure}[h!]
	\centering
	\includegraphics[scale=0.4]{images/clima.png}
	\caption{\footnotesize {Colombian weather conditions. \\}}	
	\label{fig:temp}
	{\footnotesize (a), Available photosynthetic solar radiation (PAR). (b), Daily average temperature  (c), Monthly average of precipitation (bars) and relative humidity (dotted line) from 2018 to 2021 }
\end{figure}
\newpage

@@

Calibration of crop models are conducted typically for particular cultivars and require site specific inputs of weather \citep{Crout20142}. The procedure for the SIMPLE model \citep{Zao2019simple} calibration  was a sequential process of modifying physiological variables specific for cocoa in the inputs files and  adding the appropriate files of weather for each region tested. In the SIMPLE model has seven input files where new cocoa crop data should be provided. Files in the list below can be edited to define the features of the new cultivars or experiments. Cocoa crop has not been simulated with SIMPLE model, hence consider previous cocoa studies, values such as leaf area index (LAI) \citep{Agele2016} and Harvest index (HI) \citep{Quintana2015} where modified. Radiation Use Efficiency (RUE) was calibrate according to the perennial crops  (banana and cotton) that had been calibrated previously in SIMPLE model \citep{Zao2019simple} with a RUE of 0.8 and 0.85 respectively. As cocoa trees are under shadow the RUE was lower with values between 0.7 and 0.5 g MJ$^{\mathbf{-1}}$ m$^{2}$ (table \ref{tab:reparam}).
Once the physiological parameters were calibrated to the  simulated yields for cocoa were reasonably close to the observed yield. 23 flowering dates from 12-July-2019 to 23-June-2020 were introduced in the treatment file, running the program and saving the results.

@@

@@

\begin{figure}[h!]
	\centering
	\includegraphics[scale=0.35]{images/heatm.png}\\
	\caption{\footnotesize {Pearson correlation average weather variables and flowering dates for all regions. \\}}
	\label{fig:heat} 
\end{figure}
\newpage

\subsection{Thermal time }
Thermal time characterization  was made considering 180 days after flowering (daf) for each region. The boxplot in the figure \ref{fig:ttbox} shows the data distribution where boxes indicate the range of the central 50\% of the total data per region, the central line in the box is marking the median value and lines draw out from each box mean the range of the remaining data. Therefore, this boxplot shows differences between locations as was expected following the tendencies of the temperature  (fig. \ref{fig:temp},c). Apartado and Arauca had the highest temperatures hence, the highest thermal time values with 2909 and 2764 days$^\circ$C  respectively. Caldas had the lowest values with 1173 days$^\circ$C. Meanwhile, Cali and Santander presented similar thermal time around 2000 $^\circ$C (table \ref{tab:reparam}). The accumulated temperature during the pod development (fig.\ref{fig:ttbox}) depends on the region where cocoa is cultivated and the variety planted. Thermal time values are also proportional to the yield reported on field (table \ref{tab:reparam}).\\

\begin{figure}[h!]
	\centering
	\includegraphics[scale=0.3]{images/ttbbox.png}\\
	\caption{\footnotesize {Cocoa yield and thermal time characterization at 180 days after flowering.\\}}
	\label{fig:ttbox}
\end{figure}

\subsection{Model validation}
Therefore, biomass production of the aerial part of the plant was simulated per region (fig. \ref{fig:m1}, a) as biomass-rate which is the daily biomass growth rate.  As for cocoa simulation we did not consider water and heat stress variables, in this research biomass simulation is affected by the solar radiation, RUE, daily temperature, atmospheric CO$_{2}$ concentration (ppm) and the fraction of solar radiation intercepted by a tree of cocoa during the fruit development (fSolar) (fig. \ref{fig:m1}, b). Due to the lack of this kind of data from fields, the comparison was not possible with observed data of biomass production through the crop cycle. However, the biomass simulation also responded proportionally with the yield production. Biomass of cocoa fruits simulated include all the parts that compound  the pod when they are fresh. Therefore, these this information may be of interest to farmers to predict pod fresh mass.

The figure \ref{fig:m1}, b the fSolar was simulated thought the fruit development resulting that crops in  all regions reached the maximun fSolar at 0.94, except Caldas where crops reached 0.76.  Apartado and Arauca reached faster this fSolar-max. These high values of solar radiation intercepted for photosynthesis, lasted differently depending on the region and their solar radiation (Fig.\ref{fig:temp}, a) : Apartado 66 days  from  69 to 135 days after flowering (daf),  Arauca 61 days from 72 to 133 daf, Cali 38 days from 83 to 121 daf,  Santander 24 days from 92 to 116 daf and Caldas 2 days at 125 daf. fSolar declined in the interception of solar radiation until the pod harvest day.

The cocoa yield simulation (kg ha$^{-1}$ per year) was validated using observed data (table \ref{tab:reparam}) showing a final RRMSE of 7.2\% (fig. \ref{fig:m1}, c). Individual errors per region are presented in table \ref{tab:error}, where the best fit of the calibration model  for yield prediction  was for crops in  Apartado and the highest error was calculated for Caldas crops.  The model responded to the variations of temperature and solar radiation. Therefore, the highest  yield values simulated were obtained for Arauca over 4000 kg ha$^{-1}$, followed by Apartado Santander with yields over 2000 kg ha$^{-1}$.  The lowest yield was simulated for Caldas region with less of 1000 kg ha$^{-1}$ . Final yield in the model was is calculated as the product of biomass of aerial part and harvest index (HI) \citep{Zao2019simple, Amir1991}, where the HI is similar to the CropSyst \citep{STOCKLE2003} and AquaCrop \cite{Steduto2009}.

\begin{figure}[h!]
	\centering
	\includegraphics[scale=0.4]{images/outmodel.png}
	\caption{\footnotesize {Model predictions. (a) Biomass aerial part. (b) Interception of solar radiation. (c) Yield. Crop cycle close to 180 daf (vertical red line) base on figure  \ref{fig:ttbox}.  \\ }} 
	\label{fig:m1}
\end{figure}

\begin{table}[h!]	
	\caption {\footnotesize {Summary of relative root mean square error RMSE for the predicted cocoa yield using SIMPLE model.}}
	\label{tab:error} 
	\centering
	\begin{small}
		\begin{tabular}{l c c c c c c}
			\hline
			{\bf Region }&{\bf Apartado }&{\bf Arauca}&{\bf Santander}&{\bf Cali}&{\bf Caldas}&{\bf Overal}\\
			\hline
			RMMSE \%  & 3 & 6.05 & 10.06&8.5&14.90&7.2 \\
			\hline
		\end{tabular} \\
	\end{small}
\end{table}
\newpage

\subsection{Prediction of the pod harvest day }
These result demonstrated that the traditional way to harvest always at 180 daf is not having account physiological and environmental conditions that can be affecting the pod development and maturation.
Figure \ref{fig:dayH} present that observe day of harvest or maturity day is 180 daf independently of the region. All the regions except Santander, presented the earliest harvest when the flowering date was between December and February. The fruit can be ready to harvest depending of the enviromental condition per region. Thus, when the fruit development was simulated the maturity day in Cali and Caldas were from 3 to 12 and 4 to 23 days before 180 daf (Fig. \ref{fig:dayH}, c and f respectively). Apartado presented the most similar predicted dates of harvest to 180 daf with 170 to 182 daf  (Fig. \ref{fig:dayH}, d). The pod may be harvest in Arauca (Fig.\ref{fig:dayH}, b)  between 165 and 193 daf, Santander (Fig.\ref{fig:dayH}, a) between 165 and 183 daf, ten days less than in Arauca for the same months of flowering of July and August of 2019 . Only Arauca presented longer crop cycles when the flowering was between during that period of time. This means that Arauca had bigger variation of temperature between months (Fig. \ref{fig:temp}, b ).

\begin{figure}[h!]
	\centering
	\includegraphics[scale=0.4]{images/RegionHarvest2.png}
	\caption{\footnotesize {Harvest day prediction from flowering date for (a) Santander, (b) Arauca, (c) Cali, (d) Apartado,(f) Caldas. \\ }} 
	\label{fig:dayH}
\end{figure}.
\newpage

@@

\section{Discussion}

\subsection{Weather Effects over Flower Stability and Pollination}
This study presents a new approach of SIMPLE model calibration for cocoa as tropical crop in South America for five environments. The analysis of the weather conditions (Fig. \ref{fig:heat}) did not confirmed clearly that flower are affected by the wind and rain by mechanical damage. However, farmers stated that the number of flowers pollinated decrease by months where wind and rain are high. Moreover, wind can affect the availability of tiny flies pollinators from Diptera order and from the families of of the biting midges \textit{Ceratopogonidae},  genus  \textit{Forcipomyia} \citep{Saunders1959, kaufmann1975, sotomayor2020} to reach the cocoa flowers. However, the stability of cocoa flowers is influenced by seasonal wheather conditions (abiotic) and pollination (biotic) \citep{Frimpong2014}. Therefore, pollinator population shoud be coincidence with the phenology of the flowering cocoa trees \citep{Young1983, Young2012}. Flower opening is very well synchronised between the cohorts of mature flowers opening each night \citep{Niemenak2010}. The flowers open at almost exactly the same time and rate, irrespective of their position on the trunk. Thus, unfertilised flowers abscise from the trunk approximately 1 day after flower opening  \citep{Niemenak2010}. Hence more than 90\% of unpollinated flowers fall or abscised within 32 hours after  anthesis \citep{Aneja1999}. Abscission processed of  flowers are mainly controlled  by  three  hormones: auxin, ethylene, and abscisic acid (ABA) \citep{Aneja1999}. Ethylene generally promotes abscission because it may  inhibit the transport of auxin from the leaf blade, which allow the action of ABA to promove the fall of flowers \citep{Beyer1975}. In general, environmental conditions can also  stimulate   to  a  decrease  the  auxin/ethylene relationship \citep{Aneja1999}.

In our preliminary analysis, Santanter data, showed that the number of successful flowers pollinated to produce final yield can be affected negatively mostly by the rain, TMAX and wind. Nevertheless, a better field data tracing flowers development is essential to understand if it is a mechanical or physiological effect. In contrast,  \cite{Frimpong2009, Frimpong2011} indicated that the numbers of cocoa pollinators were reduced during the dry season, but increased in the wet season. This, could be due to midges need a moist environment to develop \citep{Frimpong2014}, which is difficult during the dry season as cocoa leaves create a dried ground mat \citep{Frimpong2009, Frimpong2014}. Moreover, the lack of water during dry seasons may reduce the nutrients uptake provoking the massive flower drops \citep{Vaughton2017}.  For future studies, wind could be included in the SIMPLE model as an input to simulated this mechanical effects over number of flowers. In general, field data regarding counting flowers pollinated by month should be better reported for the region here studied.

\subsection{Thermal Time for Harvest day predictions}
We define the thermal time required to harvest cocoa pods for five Colombian regions as maturation of the fruit is related with temperature during the growth cycle \citep{lopez2018}. Previous studies have calculated thermal time for  different cocoa cultivar in Brazil and Ghana \cite{Daymond2008}. They also confirm that the fruit maturation time decrease in with an increase in temperature as was presented on others researches \citep{Alvim1974, End1991, Daymond2008}. the effects of temperature and solar radiation on fruit growth and development was previously studied by \cite{Daymond2008}, showing that crop under higher temperatures thought the crop cycle induce greater fruit losses because of physiological maturation (cherelle wilt). When the fruit is mature, seeds are able to germinate inside the pod before the harvest day \citep{lopez2018}.

Our results showed that the hotter regions such as Apartado and Arauca presented higher thermal time values (Fig. \ref{fig:ttbox} and \ref{fig:temp}). Even though, Arauca had very low values of SAR but very high temperatures, this may be caused by clouds cover. The opposite can be seen for Caldas and Santander. These, extreme relations T/SAR can compensate the crop efficiency, for example in Santander (Fig. \ref{fig:m1}).  The thermal time calculation was defined base on  180 daf because farmers cut the pods by calendar days. Previous studies stated that evaluating the level of knowledge of growers regarding cocoa crop management, showed that the harvest was in the group of activities that presented the lowest level of information by the farmers \citep{Gutierrez2020}. That is why, these results present important temperature boundaries to predict fruit maturation day. Therefore, may be other environmental factors that should be studied for further research.

\subsection{Cocoa Crop Model Simulations}
Although cocoa is a relevant crop and there is an extensive agronomic literature, there is only one physiological crop model specific for cocoa so far. The (SUCROS-Cocoa) developed by \cite{zuidema2005}. However, the code was not easy available for adaptations. In contrast, the SIMPLE model has an open code in R, which we could adapt such a model would be very useful to compare yields and predict harvest date in different climates. As the harvest day was predicted form the flowering date (Fig. \ref{fig:dayH}, consequently, biomass production and fSolar presented a crop cycle shorter than 180 daf (Fig. \ref{fig:m1}, a and b). These simulations are coincident with results presented by \cite{lopez2018}, where physiological maturity of coca pod varies from 140 to 162 daf.  Our results showed how the harvest day can vary depending on the accumulate temperature during each especific crop cycle simulated (Fig. \ref{fig:dayH}).

Biomass simulations use SIMPLE model presented similar predicted values (10000 Kg ha$^{-1}$) for coca drops in Costa Rica using SUCROScocoa \citep{zuidema2005}. Biomass simulations are a common evaluation in crop modells such as Sirius \citep{Crout20142}, SUBSTOR-potato \citep{Raymundo2017} and  DSSAT, CropSyst, STICS and WOFOST \citep{Confalonieri2016}. The approach of this research was focus  on the harvest date prediction, hence the leaves crop cycle was evaluated indirectly this study. The fraction of intercepted photosynthesis active radiation (fSolar) decreased (Fig. \ref{fig:m1}, b) when the senescence of the canopy \citep{zuidema2005}. In cocoa canopy senecence  refereed to  a group of leaves responsible  at the moment of the fruit formation to produce carbohydrates. These leaves eventually drop becoming on litter over the soil. Leaves life cycle has been simulated using crop models \citep{Crout2010, zuidema2005}, which can be the reference to improve our SIMPLE coca crop model in future studies.

Yield prediction presented a RRMSE values 7.2\%, which were significant lower than those  presented for other crops using the SIMPLE model which reported and RRMSE of 24.4\% \citep{Zao2019simple}. Resulting in reliable approach for cocoa yield prediction. In general, these results may help to improve the quality of cocoa seed considering the moment to harvest can be variable depending on weather changes.

\subsection{App development and future challenges }
The original code of SIMPLE model of \citep{Zao2019simple} was modified to make easy the implementation of this cocoa model as and app to be used in smartphones and desktops by farmers in Colombia. Therefore, the new version for cocoa crops simulation will be used to predict yield, date of harvest and biomass production, inserting only the date of flowering and region. The app development is on charge of Grupo BIOS to be deliver to farmers in Caldas initially at the end of 2021. 
This research presented and initial crop calibration that can be improved with further studies, including effects over the pod production by diseases, nutritional diffidences and abiotic stresses. The future challenge will be that traditional farmers start to harvest more aware of the environmental effects over their crops. It will be necessary that they engage growers with adapt founding from scientific studies. Moreover, It will be required the help of entrepreneurs, researchers, academics and non-specialized communities to transfers the knowledge to cocoa growers.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions}
Cocoa fruit development for harvest in the right time depend on whether conditions and principles of crop physiology and flower phenology.  This was common for the five regions.Thermal time characterization range from 1200 to 3000 days $^\circ$C, with a T$_{b}$ of 10 $^\circ$C for the fruit development.  The SIMPLE model as a generic crop model allowed to predict the harvest date with better precision than only considering days by calendar. Thus, the crop cycle of cocoa for harvest  should be shorter than 180 days after flowering. These results confirm the potential of crop model for tropical crop in Latin America.

\section{Author contributions}
ARV and ARC model calibration and data analsis. AMG field data collection. ARV were primarily responsible for writing the manuscript.

\acknowledgments{The authors wish to thank to Grupo BIOS and Fedecacao for facilitated the data from cocoa fields. }
